---
title: "Canine Confounders"
date: "2024-09-17"
bibliography: 05COVID.bib
execute: 
  echo: false
margin-footer: "[Slides](/standalone/labs/05COVID.html), [Assignment](/standalone/hw/L05COVID.html)"
---

<!-- 
GitHub Classroom:
  - Starter code: https://github.com/BIFX552-24/canine-confounders
  - Individual Assignment
  - Private
  - Grant admin access
-->


## Today's lab

-   Review of [Canine Confounders](https://livehood-my.sharepoint.com/:b:/g/personal/johnson_hood_edu/EYFlFclT0ftElomcPdau--sBrjvcqU9kQ51MVwijlHvFcw?e=7i2y1J)
-   Practice visualization using `ggplot2`

```{r}
#| message: false
# We will be using the following packages in today's lab.
library(dplyr)
library(readr)
library(stringr)
library(lubridate)
library(purrr)

library(survival)
library(ggfortify)

library(ggplot2)
library(cowplot)
theme_set(theme_cowplot())
library(patchwork)
source('https://tinyurl.com/wjm6ryj') # colorblind palettes

library(knitr)
library(kableExtra)
```

## Canine confounders

We'll be using a [COVID-19 dataset from Mexico City](https://bit.ly/3Ou4n2U) that was featured in an article in Significance Magazine titled [Canine Confounders](https://livehood-my.sharepoint.com/:b:/g/personal/johnson_hood_edu/EYFlFclT0ftElomcPdau--sBrjvcqU9kQ51MVwijlHvFcw?e=7i2y1J) [@nunez2022].

What is presented here is intentionally rough and open ended. After filling in the missing details in the [starter code](https://classroom.github.com/a/JX3MKY70), what other questions might you have? What other tests and/or visualizations might be interesting?

```{r}
# download data if we need to refresh
# note that the current link below has some changes to the data since this lesson was developed! There are at least 4 variable name changes, and the number of rows is also different. I'm pushing the current version of the small RData file for now, but be aware that there will be some updates to the code needed if we do a refresh. Go to https://bit.ly/3Ou4n2U for updated data.
# You can download the csv and RData files here:
#  csv - https://livehood-my.sharepoint.com/:x:/g/personal/johnson_hood_edu/Edg7khCfjS9NloIGLtkC1IUBw9zsCDHrSPXa5LHAPzD5oQ?e=HnbJGS
#  RData - https://livehood-my.sharepoint.com/:u:/g/personal/johnson_hood_edu/EeQzEtwzeXZLtgp3YhH8UrEBQ6O7NSngSh4f2vF7Prhysw?e=2vcAVj
if(FALSE)
{
    download.file(url = "https://archivo.datos.cdmx.gob.mx/sisver/covid19_sisver_cdmx.csv",
              destfile = "data/sisver_public.csv")
}

# only do this formatting step once
if(FALSE)
{
    sisver <- read_csv('data/sisver_public.csv') %>%
        rename(dt_symptoms = fecinisi, # date that symptoms started
               dt_admit = fechreg,     # date admitted to hospital
               dt_death = fecdef       # death date
    ) %>%
    
           # primary outcomes
    mutate(hospitalized = tipacien == 'HOSPITALIZADO',
           covid_death = !is.na(dt_death),
           intubation = intubado == 'SI',
           death_intubation = covid_death | intubation,

           # survival variables
           event_date = if_else(is.na(dt_death),   # censor at 30 days
                                dt_admit + days(30), dt_death),
           event_time = seconds_to_period(event_date - dt_admit)@day,
           t_sympt_death = Surv(event_time, covid_death),

           # risk factors and confounders
           dog = ifelse(conanima == "", NA,        # look for dogs as pets
                        str_detect(conanima, "PERR|PERO|PERRI|CANINO")),
           private = sector == "PRIVADA",          # private sector care
           pet_info = !is.na(dog),                 # has info on pets
           dog_all = ifelse(pet_info,              # add dog category for missing
                            dog, 'Missing') %>%
               factor(levels = c('Missing', 'FALSE', 'TRUE')),
           antivirals = ifelse(txantivi == "", NA, # used antivirals
                               txantivi == "SI"),
           diabetes = ifelse(diabetes == "", NA,   # has diabetes
                             diabetes == "SI"),
           hyperten = ifelse(hiperten == "", NA,   # has hypertension
                             hiperten == "SI"),
           obese = ifelse(obesidad == "", NA,      # has obesity
                          obesidad == "SI"),
           age_gt_65 = edad >= 65,                 # age > 65 years
           sex = ifelse(sexo == "MASCULINO",       # sex
                        'M', 'F')) %>%
    
    filter(dt_admit >= "2021-01-01" & dt_admit <= "2021-04-05" &
           hospitalized &
           event_time >= 0 & event_time <= 30) %>%
    
    select(dt_symptoms, dt_admit, dt_death, covid_death, event_date,
           event_time, death_intubation, t_sympt_death, dog, dog_all, 
           private, pet_info, antivirals, diabetes, hyperten, obese,
           age_gt_65, sex, sector)
    
    save(sisver, file = 'data/sisver.RData')
}else{
    load('data/sisver.RData')
}

```

:::{.content-visible when-profile="slides"}
## Canine Confounders
:::
:::{.content-visible when-profile="book"}
Let's start by looking at the primary effects of different risk factors and confounders on mortality.
:::

```{r}
table1 <- tibble(factor = c('Antivirals', 'Diabetes', 'Hyptertension',
                            'Obesity', 'Age > 65', 'Sex (male)',
                            'Private Sector', 'Dog ownership'),
                 death_intubation = list(
                     glm(death_intubation ~ antivirals, data = sisver, family = binomial),
                     glm(death_intubation ~   diabetes, data = sisver, family = binomial),
                     glm(death_intubation ~   hyperten, data = sisver, family = binomial),
                     glm(death_intubation ~      obese, data = sisver, family = binomial),
                     glm(death_intubation ~  age_gt_65, data = sisver, family = binomial),
                     glm(death_intubation ~        sex, data = sisver, family = binomial),
                     glm(death_intubation ~    private, data = sisver, family = binomial),
                     glm(death_intubation ~        dog, data = sisver, family = binomial)),
                 OR = map_dbl(death_intubation, ~ 
                                  exp(summary(.x)$coefficients[2,1])),
                 p_OR = map_dbl(death_intubation, ~ 
                                summary(.x)$coefficients[2,4]),
                   
                 death_survival = list(
                     coxph(t_sympt_death ~ antivirals, data = sisver),
                     coxph(t_sympt_death ~   diabetes, data = sisver),
                     coxph(t_sympt_death ~   hyperten, data = sisver),
                     coxph(t_sympt_death ~      obese, data = sisver),
                     coxph(t_sympt_death ~  age_gt_65, data = sisver),
                     coxph(t_sympt_death ~        sex, data = sisver),
                     coxph(t_sympt_death ~    private, data = sisver),
                     coxph(t_sympt_death ~        dog, data = sisver)
                 ),
                 HR = map_dbl(death_survival, ~
                              summary(.x)$coefficients[2]),
                 p_HR = map_dbl(death_survival, ~
                                summary(.x)$coefficients[5]))

tab <- select(table1, factor, OR, p_OR, HR, p_HR) %>%
    mutate(p_OR = format.pval(p_OR),
           p_HR = format.pval(p_HR)) %>%
    kable(digits = c(NA, 2, 4, 2, 4),
          align = c('l', 'r', 'r', 'r', 'r'))
```

:::{.content-visible when-profile="book"}
```{r}
tab
```
:::
:::{.content-visible when-profile="slides"}
```{r}
tab |>
  kable_styling(font_size = 28)
```
:::

## COVID Survival

:::{.content-visible when-profile="book"}
Here are Kaplan-Meier plots for our risk factors and confounders.
:::

```{r}
make_km_plot <- function(fact)
{
    # fit survival curve for fact
    survfit(formula(paste("t_sympt_death ~", fact)), data = sisver) %>%
        
        # plot survival curve
        autoplot(conf.int = FALSE) +
        
        # modifications to plot
        scale_color_manual(values = cbbPalette[1:2]) +
        guides(color=guide_legend(title=fact))
}

g1 <- make_km_plot('antivirals')
g2 <- make_km_plot(  'diabetes')
g3 <- make_km_plot(  'hyperten')
g4 <- make_km_plot(     'obese')
g5 <- make_km_plot( 'age_gt_65')
g6 <- make_km_plot(       'sex')
g7 <- make_km_plot(   'private')
g8 <- make_km_plot(       'dog')
```

:::{.content-visible when-profile="slides"}
```{r}
(g1 + g2 + g3 + g4) / (g5 + g6 + g7 + g8)
```
:::
:::{.content-visible when-profile="book"}
```{r}
g1
g2
g3
g4
g5
g6
g7
g8
```
:::


## Missingness

:::{.content-visible when-profile="book"}
Lets take a look at the missing rate of dog ownership by the hospital they were admitted to.
```{r}
with(sisver, table(sector, dog_all)) %>%
    kable()
```
:::
:::{.content-visible when-profile="slides"}
```{r}
with(sisver, table(sector, dog_all)) |>
    kable() |>
    kable_styling(font_size = 28)
```

## Missingness
:::
:::{.content-visible when-profile="book"}
After removing a few sectors that don't have much data, let's look at this graphically over time.
:::

```{r}
# start by removing those that have very few samples
sisver_sub <- filter(sisver, !sector %in% c('CRUZ ROJA', 'ESTATAL', 'IMSS-OPORTUNIDADES', 'SEMAR', 'UNIVERSITARIO'))

(g1 <- ggplot(sisver_sub, aes(dt_admit, fill = dog_all)) +
    geom_bar() +
    scale_fill_manual(values = cbbPalette[1:3]) +
    scale_y_log10())
```

:::{.content-visible when-profile="slides"}
## Missingness
:::

```{r}
g1 + facet_wrap(~ sector)
```


## Dog ownership and missingness by sector: Bias

:::{.content-visible when-profile="book"}
How does the risk of death change by `sector`, compared to the private sector?
:::

```{r}
mutate(sisver_sub,
       sector = factor(sector, levels = c("PRIVADA", "SSA", "IMSS", "ISSSTE", "PEMEX", "SEDENA"))) %>%
glm(formula = death_intubation ~ sector, family = binomial) %>%
    summary()
```

## References