---
title: "Population genetics simulation"
date: "2024-10-01"
execute:
  echo: true
margin-footer: "[Slides](/standalone/labs/07HWE.html), [Assignment](/standalone/hw/L07HWE.html)"
---

## Today's lab

- Review of Hardy-Weinberg equilibrium
- Population genetics simulation

```{r}
#| include: false
library(dplyr)
library(knitr)
```

## Hardy-Weinberg Equilibrium

Hardy-Weinberg equilibrium (HWE) is a principle in population genetics that describes a state of genetic stability in a population. It predicts that, under certain ideal conditions, the frequencies of genotypes are dependent on allele frequencies and will remain constant across generations.

## HWE assumptions

The Hardy-Weinberg equilibrium assumes the following:

- Large population size
- Random mating
- No mutation
- No migration
- No natural selection

:::{.content-visible when-profile="book"}
In human populations, these assumptions are rarely met, but the HWE model provides a useful baseline and is actually quite robust in practice.
:::

## HWE equation

Given two alleles, A and B, with frequencies $p$ and $q$, respectively, a population in Hardy-Weinberg equilibrium will have the following genotype frequencies:

- $p^2$ = frequency of the homozygous dominant genotype (AA)
- $2pq$ = frequency of the heterozygous genotype (AB)
- $q^2$ = frequency of the homozygous recessive genotype (BB)

:::{.content-visible when-profile="slides"}
## HWE equation
:::

The Hardy-Weinberg equation can be derived from looking at all possible matings. For example for two heterozygous parents, the probability of each genotype in the offspring is as follows:

```{r}
#| echo: false
tibble(` ` = c("A", "B"),
       A = c("P(AA) = 0.25", "P(AB) = 0.25"),
       B = c("P(AB) = 0.25", "P(BB) = 0.25")) |>
  knitr::kable()
```

:::{.content-visible when-profile="book"}
When you sum the probabilities of each genotype, conditional on the frequency of parent genotypes in the population,

$$
\sum P(genotype_{f1} | genotype_{f0}) = 1,
$$

you get the Hardy-Weinberg equation:
:::

:::{.content-visible when-profile="slides"}
## HWE equation

The Hardy-Weinberg equation is given by:
:::

$$
p^2 + 2pq + q^2 = 1
$$

## Population genetics simulation

We will use a simulation to explore Hardy-Weinberg equilibrium. To this end, we need to define the following data structures and functions:

```{r}
#' gen
#' A generation of individuals
#' @format A list with the following fields:
#' @field genos A vector of genotypes, one for each individual in the 
#'   population, with 0, 1, and 2 representing genotypes AA, AB, and BB
#' @field p The observed frequency of allele A
#' @field q The observed frequency of allele B
#' @field pAA The observed frequency of genotype AA
#' @field pAB The observed frequency of genotype AB
#' @field pBB The observed frequency of genotype BB

#' gen_list
#' A list of generations
#' @format A list of `gen` objects, one for each successive generation
```

:::{.content-visible when-profile="slides"}
## Population genetics simulation
:::

```{r}
#' gen0
#' Generate the initial population
#' @param n Number of individuals
#' @param p Frequency of allele A (frequency of B is assumed to be 1 - p)
#' @return A `gen_list` object of length 1 containing the initial generation
```

:::{.content-visible when-profile="slides"}
## Population genetics simulation
:::

```{r}
#' next_gen
#' Generate the next generation
#' @param gen_prev A `gen_list` object containing previous generations (at
#'   least one)
#' @param n Number of individuals to simulate in the next generation
#' @param p Frequency of allele A (frequency of B is assumed to be 1 - p)
#' @details Individuals of the most recent generation will be randomly 
#'   sorted and paired to generate the next generation. Offspring will be
#'   generated by randomly selecting a pair of parents, generating a single
#'   offspring, and repeating this process until the desired number of
#'   offspring is reached.
#' @return A `gen_list` object of length `length(gen_prev) + 1` including 
#'   `gen_prev` and the newly simulated generation.
```

:::{.content-visible when-profile="slides"}
## Population genetics simulation
:::

```{r}
#' plot_gen_list
#' Plot the allele and genotype frequencies of a `gen_list` object
#' @param gen_list A `gen_list` object
#' @return A ggplot object
```

## Primary tasks

* Initialize your group's repository with the provided [template](need link).

:::{.content-visible when-profile="slides"}
* Implement the functions `gen0`, `next_gen`, and `plot_gen_list` in R.

* Create a README file using Quarto with documentation.
:::

:::{.content-visible when-profile="book"}
* Implement the functions `gen0`, `next_gen`, and `plot_gen_list` in R.
  * Each student pick a different function to implement.
  * A sample `gen_list` object is included with the template.
  * Work together to test the functions.
  
* Create a README file using Quarto with documentation:
  * Use `format: "gfm"` in the YAML header.
  * Describe the use and purpose of the functions.
  * Provide examples of how to use the functions.
:::

* Commit and push your changes to the GitHub.

## Secondary tasks

* Add the following to your README file (some of these may require changes to your functions):
  * A brief description of the assumptions of the HWE model.
  * A simulation of the consequences of departures from HWE.
    * Each group will pick a different assumption to simulate.
    * Complete as many assumptions as time allows.
